experDesign: stratifyuing samples into batches with minimal bias

### 2021-07-15 paper.Rmd
Update text

---
title: 'experDesign: stratifyuing samples into batches with minimal bias'
output:
  html_document:
    df_print: paged
  md_document:
    preserve_yaml: yes
  word_document:
    keep_md: yes
date: "23 April 2021"
affiliations:
- name: Centro de Investigación Biomédica en Red, Enfermedades Hepáticas y Digestivas
  index: 1
- name: Institut d'Investigacions Biomèdiques August Pi i Sunyer, IDIBAPS
  index: 2
authors:
- name: Lluís Revilla Sancho
  orcid: 0000-0001-9747-2570
  affiliation: 1, 2
- name: Juan-José Lozano
  orcid: 0000-0001-9747-2570
  affiliation: '1'
- name: Azucena Salas
  orcid: 0000-0001-9747-2570
  affiliation: '2'
tags:
- R
- batch effect
- experiment design
bibliography: paper.bib
---

```{r setup}
knitr::opts_chunk$set(collapse = TRUE)
```

@@

Randomization and replicates can be utilized to prevent batch effects to confound the analysis.
However, by examining how the variables are distributed across each batch sample, proper randomization can be ensured, thus minimizing batch effects.
This is known as randomized block experimental design or stratified random sampling experimental design.

@@

There are some tools to prevent batch effects on the R language in multiple fields and areas, and particularly for biological research [@rcoreteam2014].
Here we briefly describe the currently available packages:

@@

```{r Omixer, error=TRUE}
library("Omixer")
data(survey, package = "MASS")
survey$ID <- seq_len(nrow(survey))
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, 
                           plateNum = n_batch, randVars = VoI, techVars = NULL)
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, mask = rep(1, nrow(survey)),
                           plateNum = n_batch, randVars = VoI, techVars = NULL)
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, mask = rep(0, nrow(survey)),
                           plateNum = n_batch, randVars = VoI, techVars = NULL)
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, mask = rep(1, 288),
                           plateNum = n_batch, randVars = VoI, techVars = NULL)
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, mask = rep(0, 288),
                           plateNum = n_batch, randVars = VoI, techVars = NULL)

@@

## experDesign

***experDesign***, the package described on this paper, first evaluates the distribution of samples in batches and then on a spatial grid.

@@

The package ***experDesign*** provides the functional `design` to distribute the samples into multiple batches such that each variable is homogeneous within each batch.
Each batch is set to have some centrality and dispersion statistics to match as closely as possible the original input design data.
It does by comparing the mean, the standard deviation, the median absolute deviation, variables with no value number, the entropy and the independence of the categorical variables.
The distribution of samples that more closely match the original distribution is returned to the user.

This approach is similar to the *anticlust* method used for maximum variance, where the objective function is the sum of the squared errors between cluster centers and individual data points.
But instead of comparing to the clusters center it is compared to the whole data.
If the experiment is carried out with a specific spatial distribution, the `spatial` function also distributes the samples homogeneously by position.

@@

```{r controls}
extreme_cases(survey[, VoI], size = 5)
replicates_index <- replicates(survey[, VoI], size_subset = size_subset, controls = 5, iterations = iterations)
spatial_replicate_index <- spatial(replicates_index, survey[, VoI])
inspect(spatial_replicate_index, survey[, VoI])
```

***experDesign*** also provides several small utilities to make it easier to design the experiment in batches.
For instance, a function called `sizes_batches` helps calculate the number of samples in order to distribute them across the number of batches required.

@@

>>>>

---
title: 'experDesign: stratifyuing samples into batches with minimal bias'
output:
  word_document:
    keep_md: yes
  md_document:
    preserve_yaml: yes
  pdf_document: default
  html_document:
    df_print: paged
date: "15th July 2021"
affiliations:
- name: Centro de Investigación Biomédica en Red, Enfermedades Hepáticas y Digestivas
  index: 1
- name: Institut d'Investigacions Biomèdiques August Pi i Sunyer, IDIBAPS
  index: 2
authors:
- name: Lluís Revilla Sancho
  orcid: 0000-0001-9747-2570
  affiliation: 1, 2
- name: Juan-José Lozano
  orcid: 0000-0001-9747-2570
  affiliation: '1'
- name: Azucena Salas
  orcid: 0000-0001-9747-2570
  affiliation: '2'
tags:
- R
- batch effect
- experiment design
bibliography: paper.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, warning = TRUE, error = TRUE)
```

@@

Randomization and replicates can be used to prevent batch effects to confound the analysis.
However, by examining how the variables are distributed across each batch sample, proper randomization can be ensured, thus minimizing batch effects.
This is known as randomized block experimental design or stratified random sampling experimental design.

@@

There are some tools to minimize batch effects on the R language in multiple fields and areas, and particularly for biological research [@rcoreteam2014].
Here we briefly describe the currently available packages:

@@

```{r Omixer, error=TRUE}
library("Omixer")
survey$ID <- seq_len(nrow(survey))
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, 
                           plateNum = n_batch, randVars = VoI, techVars = NULL)
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, mask = rep(1, nrow(survey)),
                           plateNum = n_batch, randVars = VoI, techVars = NULL)
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, mask = rep(0, nrow(survey)),
                           plateNum = n_batch, randVars = VoI, techVars = NULL)
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, mask = rep(1, 288),
                           plateNum = n_batch, randVars = VoI, techVars = NULL)
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, mask = rep(0, 288),
                           plateNum = n_batch, randVars = VoI, techVars = NULL)

@@

The package ***experDesign*** provides the functional `design` to distribute the samples into multiple batches such that each variable is homogeneous within each batch.
Each batch is set to have some centrality and dispersion statistics to match as closely as possible the original input design data.
It does by comparing the mean, the standard deviation, the median absolute deviation, variables with no value number, the entropy and the independence of the categorical variables.
On each iteration if the distribution of samples for each batch has less differences with the original distribution than the last previous stored sample distribution then it replace it as the best sample distribution.
At the end of the iterations the best sample distribution is returned to the user.

@@

One can check the statistics of the index for multiple statistics:

```{r evaluate}
evaluate_na(experDesign_index, survey[, VoI])
evaluate_entropy(experDesign_index, survey[, VoI])
evaluate_mad(experDesign_index, survey[, VoI])
evaluate_sd(experDesign_index, survey[, VoI])
evaluate_mean(experDesign_index, survey[, VoI])
# All together for each batch
ei <- evaluate_index(experDesign_index, survey[, VoI])
ei[, , "SubSet1"]
evaluate_na(experDesign_index_spatial, survey[, VoI])
evaluate_entropy(experDesign_index_spatial, survey[, VoI])
evaluate_mad(experDesign_index_spatial, survey[, VoI])
evaluate_sd(experDesign_index_spatial, survey[, VoI])
evaluate_mean(experDesign_index_spatial, survey[, VoI])
# All together for each batch
ei <- evaluate_index(experDesign_index_spatial, survey[, VoI])
ei[, , "G1"]
```

And compare it with the original distribution:

```{r original}
evaluate_orig(survey[, VoI])
```

In ***experDesign*** the difference is used to compare with the original distribution, this approach is similar to the *anticlust* method used for maximum variance, but instead of the difference with the original distribution the objective function is the sum of the squared errors between cluster centers and individual data points.
But instead of comparing to the clusters center it is compared to the whole data.
If the experiment is carried out with a specific spatial distribution, the `spatial` function also distributes the samples homogeneously by position.
See below for an example.

@@

```{r controls}
extreme_cases(survey[, VoI], size = 5)
replicates_index <- replicates(pheno = survey[, VoI], 
                               size_subset = size_subset, controls = 5, 
                               iterations = iterations)
spatial_replicate_index <- spatial(index = replicates_index, pheno = survey[, VoI], 
                                   rows = LETTERS[1:8], columns = 1:12)
survey$ID <- seq_len(nrow(survey))
report_replicates <- inspect(i = replicates_index, 
                             pheno = survey[, c(VoI, "ID")])
report_replicates_position <- inspect(i = spatial_replicate_index, 
                             pheno = report_replicates, 
                             index_name = "position")
# Batch and position where the samples should be:
head(report_replicates_position)
# Technical replicates:
head(report_replicates_position[duplicated(report_replicates_position$ID), ])
```

***experDesign*** also provides several small utilities to make it easier to design the experiment in batches.
For instance, a function called `sizes_batches` helps calculate the number of samples in order to distribute them across the number of batches required and another one calculates the minimal amount of batches required:

@@

In conclusion the ***experDesign*** can be a first fast step to prepare for a batch experiment.
It can use as many numeric and categorical variables as needed to stratify the experiment design on batches including spatial distributions.
